### 2017-01-29 manuscript.rmd
typo

Functional genomics, the study of how genomes work, is one of the most rapidly advancing fields in biology. There is now considerable interest in comparing genome function across species to understand what is conserved, what is variable, and how the diversity of organism traits (including anatomy, physiology, and development) relates to genome diversity. Most published comparisons of genome function across species have relied on multiple pairwise comparisons between species, an approach that neglects to account for the evolutionary relationships among species. This has two potential negative impacts. First, it can lead to the wrong conclusions about the evolution of genome function. Second, it is a missed opportunity to learn about biology that can only be understood in an explicit historical context that considers these relationships. Here we examine two recently published comparative gene expression studies that used multiple pairwise comparison rather than phylogenetic approaches to investigate evolutionary patterns of gene expression. We find problems with the pairwise comparisons used in both studies that undermine their published conclusions. The first study we examine found that expression tends to be more different between paralogs than orthologs, and interpreted this as evidence of the ortholog conjecture- the hypothesis that there are greater rates of evolutionary change following duplication. Our phylogenetic analysis finds no difference in rates following duplication and speciation. Instead, the greater difference between paralogs is because they tend to be more distantly related to each other than orthologs are to each other. The second study claimed to find more evolutionary variance in gene expression at a midpoint of embryonic development than earlier and later when comparing distantly related animals. We instead find that changes along the branch of a single taxon, the ctenophore, were over counted because they were included in each pairwise comparison with this species. These reanalyses are concrete demonstrations of the inadequacy of pairwise comparisons in comparative genomics studies, including gene expression analyses, and indicates that it will be critical to adopt phylogenetic comparative methods in future work.

>>>>

Functional genomics, the study of how genomes work, is one of the most rapidly advancing fields in biology. There is now considerable interest in comparing genome function across species to understand what is conserved, what is variable, and how the diversity of organism traits (including anatomy, physiology, and development) relates to genome diversity. Most published comparisons of genome function across species have relied on multiple pairwise comparisons between species, an approach that neglects to account for the evolutionary relationships among species. This has two potential negative impacts. First, it can lead to the wrong conclusions about the evolution of genome function. Second, it is a missed opportunity to learn about biology that can only be understood in an explicit historical context that considers these relationships. Here we examine two recently published comparative gene expression studies that used multiple pairwise comparisons rather than phylogenetic approaches to investigate evolutionary patterns of gene expression. We find problems with the pairwise comparisons used in both studies that undermine their published conclusions. The first study we examine found that expression tends to be more different between paralogs than orthologs, and interpreted this as evidence of the ortholog conjecture- the hypothesis that there are greater rates of evolutionary change following duplication. Our phylogenetic analysis finds no difference in rates following duplication and speciation. Instead, the greater difference between paralogs is because they tend to be more distantly related to each other than orthologs are to each other. The second study claimed to find more evolutionary variance in gene expression at a midpoint of embryonic development than earlier and later when comparing distantly related animals. We instead find that changes along the branch of a single taxon, the ctenophore, were over counted because they were included in each pairwise comparison with this species. These reanalyses are concrete demonstrations of the inadequacy of pairwise comparisons in comparative genomics studies, including gene expression analyses, and indicates that it will be critical to adopt phylogenetic comparative methods in future work.
